Tree Models

### 2025-12-10 Presentation DSCI445.Rmd
Updates to the Presentation

---
title: "DSCI 445 Presentation"
author: "Ethan Schilling, Joshua Tobey, Dawson Carney, Reece Carmody"
date: "2025-12-04"
output: beamer_presentation
---

@@

## Slide 1

```{r}
readRDS("TOC_plot.rds")
readRDS("TDS_plot.rds")
```

>>>>

---
title: "DSCI 445 Presentation"
author: "Ethan Schilling, Joshua Tobey, Dawson Carney, Reece Carmody"
date: "2025-12-11"
output: 
  beamer_presentation:
    theme: "Warsaw"
    colortheme: "default"
---

@@

```{r, include=F}
library(dplyr)
library(tidyr)
library(readr)
library(lubridate)
library(ggplot2)
library(zoo)
library(purrr)
library(knitr)
library(patchwork)
library(tidyverse)
library(tidymodels)
library(rpart.plot)
library(kableExtra)
library(vip)
```

# Example Slide Title

Example text.

- Bullet points
- Bullet points

## New section of text

More example text

```{r}
#readRDS("TOC_plot.rds")
#readRDS("TDS_plot.rds")
```

# Background and Motivation

The drinking water treatment process takes water from a river, lake, reservoir, or other source, and purifies it to have it reach drinking water standards. There are four primary steps of the water treatment process:

1. Coagulation - Chemicals ("coagulants") are added to raw water to help contaminants
group together into "floc" particles  
2. Flocculation - The water is then slowly mixed to allow these floc particles to grow  
3. Filtration - These particles are filtered out  
4. Disinfection - Water is disinfected to get rid of biological contaminants

This project will focus on Step 1. These chemical doses are one of the most important
features of the treatment process that an operator can change to improve performance.

# Background and Motivation

**The role of an operator in deciding water treatment plant chemical doses is to
look at a wide array of factors and make decisions about how to adjust
chemical doses.**

- Treatment plant conditions change rapidly
- A predictive tool to help select chemical doses is beneficial both financially and environmentally

```{r}
# Add explanation picture/plot to help people visualize here??
```

# Data Overview

Three years of data from 2018-2020, provided by a water treatment plant in Aurora, Colorado.

- Raw Water Data
   - pH
   - Temperature (of the water)
   - Turbidity
   - Conductivity
   - Suspended Grain Size
   - Alkalinity
   - Hardness

- Chemical Dosing Data
   - Ferric Chloride Dose: primary additive
   - Cationic Polymer Dose: secondary additive

# Data Cleaning

- Features were averaged to match the dosing data's daily increments 
- Merged various datasets into a final cleaned dataset
- Various outliers (likely caused by equipment malfunction) were either removed or interpolated
- Suspended Grain Size was not used as it was highly variable
- Alkalinity and Hardness was not used because of missing data chunks and differing time ranges

# Exploratory Analysis

```{r, fig.width=7, fig.height=2}
df_lag_incdec <- read.csv("RWAR_dosing_lag_incdec.csv") |>
  mutate(Timestamp = ymd(Timestamp))

plot_polymer <- ggplot(df_lag_incdec, aes(x = Timestamp,
                           y = Polymer_dose,
                           color = Polymer_dose)) +
  geom_line() + 
  scale_color_gradient(low = 'yellow', high = 'green') +
  theme_minimal() +
  ggtitle("Polymer Coagulant")

print(plot_polymer)

plot_FeCl <- ggplot(df_lag_incdec, aes(x = Timestamp,
                           y = FeCl_dose,
                           color = FeCl_dose)) +
  geom_line() + 
  scale_color_gradient(low = 'blue', high = 'orange') +
  theme_minimal() +
  ggtitle("FeCl Coagulant")

print(plot_FeCl)
```

# Exploratory Analysis

```{r, fig.width=7, fig.height=2}

plot_pH <- ggplot(df_lag_incdec, aes(x = Timestamp,
                           y = mean_pH,
                           color = mean_pH)) +
  geom_line() + 
  scale_color_gradient(low = 'blue', high = 'red') +
  theme_minimal() +
  ggtitle("Average pH")

print(plot_pH)

plot_temp <- ggplot(df_lag_incdec, aes(x = Timestamp,
                           y = mean_Temp,
                           color = mean_Temp)) +
  geom_line() + 
  scale_color_gradient(low = 'blue', high = 'red') +
  theme_minimal() +
  ggtitle("Water Temperature")

print(plot_temp)

```

# Exploratory Analysis

```{r, fig.width=7, fig.height=2}

plot_cond <- ggplot(df_lag_incdec, aes(x = Timestamp,
                           y = mean_Conductivity,
                           color = mean_Conductivity)) +
  geom_line() + 
  scale_color_gradient(low = 'purple', high = 'yellow') +
  theme_minimal() +
  ggtitle("Water Conductivity")

print(plot_cond)

plot_turb <- ggplot(df_lag_incdec, aes(x = Timestamp,
                           y = mean_Turbidity,
                           color = mean_Turbidity)) +
  geom_line() + 
  scale_color_gradient(low = 'black', high = 'orange') +
  theme_minimal() +
  ggtitle("Water Turbidity")

print(plot_turb)

```

# Predictor Selection

Generally, temperature and pH seem to have the strongest relationship. This coincides with feedback from treatment plant operators. For further analysis, we will use:

- pH
- Temperature
- Conductivity
- Turbidity

# Introducing Lagged Variables

Similar to the Stock dataset we've seen in labs and homeworks, our time series dataset needs lagged variables.

- For each of the main four parameters:
   - Introduce two lagged variables
   - First represents the day before
   - Second represents two days before

# Coagulant Dose Prediction

**Add slides for various numerical prediction models**

# Coagulant Dose Prediction: Decision Tree

```{r, fig.width=7, fig.height=3.8}
set.seed(445)

# Decision Tree

# For now just predicting numerical amounts, not classifying
data = read.csv("RWAR_dosing_lag_incdec.csv")
data = as.data.frame(data)
data$FeCl_change = NULL
data$Polymer_change = NULL

# Set up training & testing data

# Can change these if you want to modify the tree/data
tree_depth = 5 # tree depth
min_n = 50 # minimum observations needed to further split

train_data = data %>%
  filter(year(Timestamp) <= 2019)
test_data = data %>%
  filter(year(Timestamp) >= 2020)

# Create the tree

recipe = recipe(FeCl_dose ~ ., data = train_data) %>% # Change target variable here
  step_rm(Timestamp) %>%
  step_rm(Polymer_dose) # Remove the other target variable here

tree_spec = decision_tree(tree_depth = tree_depth, min_n = min_n) %>%
  set_engine("rpart") %>%
  set_mode("regression")

workflow = workflow() %>%
  add_recipe(recipe) %>%
  add_model(tree_spec)

fit_tree = workflow %>%
  fit(data = train_data)

# Plot the tree

fit_tree %>%
  extract_fit_engine() %>%
  rpart.plot(roundint = FALSE)

# Assess accuracy

predictions = fit_tree %>%
  predict(test_data) %>%
  bind_cols(test_data)

actual = predictions$`FeCl_dose` # Change target variable here
predicted = predictions$.pred

rmse = sqrt(mean((actual - predicted)^2))
r_squared = cor(actual, predicted)^2
mae = mean(abs(actual-predicted))

metrics = data.frame(
  Metric = c("RMSE", "R-Squared", "MAE"),
  Value  = c(rmse, r_squared, mae)
)

kable(metrics, booktabs = TRUE, digits = 3) %>%
  kable_styling(position = "center", font_size = 12)

```

# Coagulant Dose Prediction: Random Forest

**Note - make a cleaner table!**

```{r, fig.width=7, fig.height=3.8}
set.seed(445)

# Random Forest

# For now just predicting numerical amounts, not classifying
data = read.csv("RWAR_dosing_lag_incdec.csv")
data = as.data.frame(data)
data$FeCl_change = NULL
data$Polymer_change = NULL

# Set up training & testing data

# Can change these if you want to modify the tree/data
trees = 500 # number of trees in the forest
min_n = 50 # minimum observations needed to further split

train_data = data %>%
  filter(year(Timestamp) <= 2019)
test_data = data %>%
  filter(year(Timestamp) >= 2020)

# Create the tree

recipe = recipe(FeCl_dose ~ ., data = train_data) %>% # Change target variable here
  step_rm(Timestamp) %>%
  step_rm(Polymer_dose) # Remove the other target variable here

rf_spec = rand_forest(trees = trees, min_n = min_n) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

workflow = workflow() %>%
  add_recipe(recipe) %>%
  add_model(rf_spec)

fit_rf = workflow %>%
  fit(data = train_data)

# Assess accuracy

predictions = fit_rf %>%
  predict(test_data) %>%
  bind_cols(test_data)

actual = predictions$`FeCl_dose` # Change target variable here
predicted = predictions$.pred

# Metrics and plots

model_metrics = predictions %>%
  metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here

print(model_metrics)

vip(fit_rf, bar = TRUE) +
  theme_minimal(base_size = 14) +
  labs(title = "Top Feature Importances")

```

# Coagulant Change Prediction

**Add slides for various classification models**

### 2025-12-10 Tree Models Reece C.Rmd

library(dplyr)
library(tidyverse)
library(tidymodels)

@@

# Can change these if you want to modify the tree/data
tree_depth = 5 # tree depth
min_n = 50 # minimum observations needed to further split
train_proportion = 0.8 # split of data

n = nrow(data)
indices = sample(1:n, size = floor(train_proportion * n))

train_data = data[indices, ]
test_data  = data[-indices, ]

@@

rmse = sqrt(mean((actual - predicted)^2))
print(paste("RMSE: ", rmse))

r_squared = cor(actual, predicted)^2
print(paste("R Squared: ", r_squared))

@@

library(rpart.plot)

@@

library(dplyr)
library(tidyverse)
library(tidymodels)

@@

# Can change these if you want to modify the tree/data
trees = 500 # number of trees in the forest
min_n = 50 # minimum observations needed to further split
train_proportion = 0.8 # split of data

n = nrow(data)
indices = sample(1:n, size = floor(train_proportion * n))

train_data = data[indices, ]
test_data  = data[-indices, ]

@@

rf_spec = rand_forest(trees = trees, min_n = min_n) %>%
  set_engine("ranger") %>%
  set_mode("regression")

@@

rmse = sqrt(mean((actual - predicted)^2))
print(paste("RMSE: ", rmse))

r_squared = cor(actual, predicted)^2
print(paste("R Squared: ", r_squared))

>>>>

library(dplyr)
library(tidyverse)
library(tidymodels)
library(rpart.plot)
library(knitr)
library(kableExtra)

@@

# Can change these if you want to modify the tree/data
tree_depth = 5 # tree depth
min_n = 50 # minimum observations needed to further split

train_data = data %>%
  filter(year(Timestamp) <= 2019)
test_data = data %>%
  filter(year(Timestamp) >= 2020)

@@

rmse = sqrt(mean((actual - predicted)^2))
r_squared = cor(actual, predicted)^2
mae = mean(abs(actual-predicted))

metrics = data.frame(
  Metric = c("RMSE", "R-Squared", "MAE"),
  Value  = c(rmse, r_squared, mae)
)

kable(metrics, booktabs = TRUE, digits = 3) %>%
  kable_styling(position = "center", font_size = 12)

@@

@@

library(dplyr)
library(tidyverse)
library(tidymodels)
#library(yardstick)
library(vip)

@@

# Can change these if you want to modify the tree/data
trees = 500 # number of trees in the forest
min_n = 50 # minimum observations needed to further split

train_data = data %>%
  filter(year(Timestamp) <= 2019)
test_data = data %>%
  filter(year(Timestamp) >= 2020)

@@

rf_spec = rand_forest(trees = trees, min_n = min_n) %>%
  set_engine("ranger", importance = "impurity") %>%
  set_mode("regression")

@@

# Metrics and plots

model_metrics = predictions %>%
  metrics(truth = FeCl_dose, estimate = .pred) # Change target variable here

print(model_metrics)

ggplot(predictions, aes(x = FeCl_dose, y = .pred)) +
  geom_point() +
  geom_abline(intercept = 0, slope = 1) +
  labs(title = "Random Forest Accuracy",
       x = "Actual",
       y = "Predicted") +
  theme_bw()

vip(fit_rf)
